Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-11-30 ResponseToReviewers.txt

Comments from Reviewer #2:
--------------------------------
Authors propose a new theory describing epidemic spreads, which is based on the hypothesis of the action of an environmental stress, with a proposal of unification of SIR model and Gompertz model. To this end they resort to a microscopic (networked) description in which the macroscopic variable derives from a geometric mean instead from an arithmetic one. In their approach the usual set of
equations in a networked model is replaced by one in which the causality between the infected and susceptible populations is reversed.

First of all, the usual networked equation for the node i is

d x_i (t)/dt = \beta (1 - x_i (t)) \sum_{j} a_{ij) x_j (t) - \gamma x_i (t),    \beta, \gamma>0,

while in the article the last term is omitted. Is this a motivated choice or a misprint?

Anyway, authors replace these equations by the set

d x_i (t)/dt = \beta x_i (t)) \sum_{j} a_{ij) (1 - x_j (t)).

To understand meaning and plausibility of the equations one can consider a single node in very simplified situations.

In the first usual form, if a_{ij) is not zero only for j different from i and the x_j (j different from i) are all zero (connected sites not infected), then x_i goes exponentially to zero also if it starts from an initial non zero value. In the same hypothesis on a_{ij}, for \gamma = 0 and x_j = 1 for all j different from i (all the connected site surely infected) x_i goes exponetially to 1 for large times. These behaviors appear reasonable.

Let us look now to the form proposed by the authors. In the same hypothesis that a_{ij} is not zero only for j different from i, if the x_j (j different from i) are all zero (connected sites not infected) the probability x_i of the site i to be infected goes to infinity (unless \beta < 0), while if the x_j (j different from i) are all 1 (all connected sites surely infected) the probability x_i keeps its initial value at all times. This seams not reasonable.

Therefore, the conceptual hypotheses of the autors could be interesting, but they have been traduced in not sound arguments.

In conclusion, the article is not suitable for publication.

Response to reviewer #2:
--------------------------------
The reviewer makes one specific point about the microscopic logistic and modified logistic model and one the basis of this comment deems our publiation not suitable for publication. Let us first address the specific critisicm, which related to the plausibility of the modified SIR model, where causality is reversed, meaning that the "susceptible" drives the "infections". Because we are operating under a new paradigm, we do not consider an infected person as infectious, but rather consider the susceptible as the drivers of the growth, as they represent the stage of the environmental stressor. For the case where all other entities are infected (x_j = 1), the initial value equals the final value. This is because there are no more susceptible entities which is interpreted as the environmental stressor having passed. For the case where all other entities are susceptible (x_j = 0), the growth rate will be at it's maximum, exactly as in the gompertz model, meaning that we are just at the start of the environmental stressor. So the equations do indeed reflect the desired outcome, but they must be taken under the correct paradigm. Consequently, they would appear nonsensical if applied under the opposite paradigm. The text has now included a more detailed discussion on both of these simple yet instructive scenarios, and we thank the reviewer for pointing out this confusion.

Further, the reviewer points out that the SIR equation for infected was wrongly quoted. We see that the confusion arose because the section was titled "Unifying the SIR with the Gompertz" and we thus proceeded to start with the logistic equation *not* the SIR infection equation, hence the confusion. However, to make it more clear, the section has been renamed to "Comparing logistic and Gompertzian growth in the microscopic domain". We then proceed with the actual unification in the proceeding section now named "Unifying the SIR model with Gompertzian growth". Hopefully this will be more clear.

Response to reviewer #3:
--------------------------------
Reviewer #3: After reading the article reviewed, I came to the following conclusions: from a mathematical point of view, the model seems to be correct (unless I have overlooked a mistake), while the interpretation of the results obtained is, to put it mildly, controversial, and in some respects incorrect or requiring additional explanations. In particular:

@@

"
In conclusion, we have shown that Gompertzian growth follows from infinite interactions between the susceptible and infected states. In this vein, the Richards parameter, $\theta$, is related although not identical to the number of higher-order interactions with the susceptible and the infected population in an augmented SIR model \cite{richards1959flexible}, where infinite interactions corresponds to $\theta\rightarrow 0$ and where the source of growth is driven by the number of susceptible and not the number of infected.
[..]
Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth. 
"

Additional remarks: A comparison of the SIRD with Gompertz model, graphically presented in Fig.1 does not reflect the differences of the models considered in an acceptable way. In order to compare the models, I suggest to calculate the basic reproduction number R0, which indicates how many secondary infections are caused by an infected individual at the beginning of an epidemic when infection spreading is of exponential type. Indispensable in the calculations formulae and data one may find in the Kong et al. work (PLoS One 16, e0252373, 2021). References [12], [48], [51] are incomplete and should be supplemented with the missing information.

Answer:
Thank you for these suggestions. We have now modified our calculations to quote the basic reproduction number as modeled by Carletti et al. (see references) instead of the transmission rate. We also use a bootstrap to get reasonable confidence intervals (unlike Carletti).

### 2022-11-30 arxiv/main.tex
finalize reviewer two response

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame? While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model. This alternative theory represents an alternative paradigm in disease spreading dynamics. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

@@

Without this approximation, the full 3-parameters in Eq. \ref{eq:SIRD_D} of the deceased group can show a curve quite close to a straight line in the double-log-domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters \cite{roda2020difficult}.

@@

\section*{Unifying the SIR model with Gompertzian growth}
We now reconcile the SIR model with the Gompertz curve. 
Start with the networked SIR equation for the infected group,
\begin{equation}
\label{eq:SIRnetwork}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i,
\end{equation}

@@

Thus, Letting $s_i = 1 - x_i$ be the probability of being susceptible,

@@

To build some intuition, we consider some simple scenarios. 
\begin{enumerate}
\item Let $x_j=1$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the environmental stressor is gone, and the value of $x_i$ goes to zero. In the communicable disease paradigm, Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will growth as long as the right-hand side is positive.
\item Let $x_j \rightarrow 0$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases. This behavior is in line with Gompertzian growth and is expected as growth is driven by the susceptible. The implication is that the environmental stressor has just started. In the communicable disease paradigm Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
\end{enumerate}
It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq. \ref{eq:medianGomp},
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}
We see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and correlated to each other as under an environmental stressor.

Finally we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model. Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)}.
 \end{equation}
 Note that for $N=1$, this relative growth rate becomes the standard logistic function, while higher order terms will move closer to the Gompertz curve as the hypergerometric term approaches zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}).

@@

Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected, and be driven by the susceptible as opposed to the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.

We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems, borrowing the definition of coherence from quantum physics \cite{molski2003coherent}.

Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles. The emergence of coherent quantum phenomena at the macroscopic level suggests that no longer can the microscopic world claim a monopoly on quantum physics, especially as it relates to biology \cite{lambert2013quantum}. One might be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.

@@

The transmission rate is modeled as the product of $\eta \xi$ in Eq. \ref{eq:SIRD_D}, and its associated variance is obtained assuming a bi-normal relationship between the two parameters with cross-correlation $\rho$ \cite{nadarajah2016distribution},
\begin{equation}
\text{Var}[\beta] = \text{Var}[\eta] \text{Var}[\xi] (1 + \rho^2) + \text{Var}[\xi]*\text{E}[\eta]^2 + \text{Var}[\eta] \text{E}[\xi]^2,
\end{equation}
where estimates are obtained by substituting the model parameters with those estimated through the technique outlined above.

>>>>

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,liu2020emerging,medica2020covid,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame? While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model. This alternative theory represents an alternative paradigm in disease spreading dynamics. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

@@

Without this approximation, the full 3-parameters in Eq. \ref{eq:SIRD_D} of the deceased group can show a curve quite close to a straight line in the double-log-domain even in the initial observations, although there will always be a non-zero concavity (Fig. \ref{fig:Gompertz_vs_SIRD_lnln_infty} and the SI Appendix).
Meanwhile, even though the Gompertz model can be fit with a 3-parameter model as shown in Eq. \ref{eq:GompertzODE}, it can also be simplified to a 2-parameter model estimated through linear regression of log-mortality \cite{Levitt2020},
\begin{equation}
\label{eq:GompertzODE2param}
\frac{d}{dt}\ln{Y(t)} = -\beta\left[\ln{Y(t)} - \ln {Y_{\infty}}\right].
\end{equation}
Thus, through linear regression estimates, the Gompertz model mitigates the possibility of non-identifiability issues of the parameters, which plagues more complex models \cite{roda2020difficult}.

@@

\section*{Unifying the SIR model with Gompertzian growth}
It is possible to reconcile the SIR model with the Gompertz curve under a common model different from the traditional Richards model. First, start with the networked SIR equation for the infected group,
\begin{equation}
\label{eq:SIRnetwork}
\frac{d x_i}{dt} = \beta (1-x_i)\sum_j{a_{ij}}x_j - \alpha x_i\quad \forall i.
\end{equation}

@@

Thus, letting $s_i = 1 - x_i$ be the probability of being susceptible,

@@

To build some intuition about these equations, we consider some simple scenarios. 
\begin{enumerate}
\item Let $x_j=1$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} becomes $\frac{d x_i}{dt} = - \alpha x_i$, implying as expected that the environmental stressor is gone, and the value of $x_i$ goes to zero. In the communicable disease paradigm, Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = \beta (1-x_i) - \alpha x_i$, and $x_i$ will grow as long as the right-hand side is positive. This is expected as all neighbors are able to transmit the disease.
\item Let $x_j \rightarrow 0$, $\forall j \neq i$. In this scenario, Eq. \ref{eq:modSIR} $\frac{1}{x_i}\frac{d x_i}{dt} \rightarrow \infty$, implying asymptotically infinite initial growth as the number of higher order terms increases. This behavior is in line with Gompertzian growth and is expected as growth rates are indicated by the susceptible. The implication is that the environmental stressor has just started. In the communicable disease paradigm Eq. \ref{eq:SIRnetwork} becomes $\frac{d x_i}{dt} = - \alpha x_i$, and the value of $x_i$ goes to zero due to the absence of infectious people.
\end{enumerate}
It is important to note that both sets of equations only describe the paradigm they represent. Consequently, the set of equations for one paradigm will appear nonsensical in the other paradigm.

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log-domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha$ approaching $0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain Eq. \ref{eq:medianGomp},
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}
We thus see that in a mathematically well-defined scenario, nodes in the network respond instantaneously and are correlated with each other as under an environmental stressor.

Finally, we are positioned to arrive at a closed-form expression for a finite number of interaction terms in this modified logistic model, which represents the infection term in the SIR model. Working with a single unitless growth variable and omitting the multivariate network without loss of generality, we start with

\begin{equation}
\frac{d \ln{x}}{dt} = \sum_{i=1}^N s^i/i.
\end{equation}
Using standard integrals we see that the multiple terms amount to a Gompertz growth term adjusted by a Hypergeometric function,
\begin{eqnarray*}
  \sum_{i=1}^N s^i/i &=& \int_s \sum_{i=0}^{N-1} s^i ds\\
  &=&\int_s \frac{1-s^N}{1-s} ds\\
  &=&- \frac{{}_{1}F_{2}({1,N+1,N+2;s})}{N+1}s^{N+1} - \ln(1-s),
\end{eqnarray*}
where ${}_1F_{2}$ is Gauss' Hypergeometric function, and integration limits are indefinite. In summary, the modified logistic function emerges as
 \begin{equation}
 \label{eq:modLogistic}
\frac{d}{dt}\ln{x(t)} = - \frac{{}_{1}F_{2}({1,N+1,N+2;1-x(t)})}{N+1}(1-x(t))^{N+1} - \ln{x(t)},
 \end{equation}
 which represents the novel bridge between logistic and Gompertzian growth. Note that for $N=1$, this relative growth rate becomes the standard logistic function, while increasing the number of higher order terms will move the equation closer to the Gompertz curve as the Hypergeometric term approaches zero. This modification to the logistic function does not have a 1-to-1 correspondence with the Richards model, but serves instead as an alternative when encoding non-linear or collaborative growth effects (see Fig. \ref{fig:hypergeometric}). This novel growth equation can thus be included in the SIR equation for the infected to model collaborative growth.

@@

Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.

We further show that the observed mortality across countries can be explained by a model where the biosphere is stressed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. The stressor is modeled as a stochastic perturbation in the log-transformed domain of effects where correlation between people or microscopic entities is present at all orders. From this model, we draw parallels between the coherent behavior of the population's mortality evolution during an epidemic and the spatial coherence of quantum mechanical systems.

Thus, we see growth on a spectrum: In one extreme we find non-collaborative growth models or models with parameterized linear interaction effects, and in the other extreme we see a field of microscopic entities coherently sharing information much like quantum entangled particles. This observation adds to the list of emerging biological phenomena that exhibit coherent quantum behavior at the macroscopic level \cite{lambert2013quantum}. Although en environmental stressor can explain our observations, one might still be surprised to find that temporal evolution of human mortality during epidemics can behave like the spatial energy distribution of quantum coherent systems.

@@

The basic reproduction number is modeled as the product of $\eta \xi / \kappa$ in Eq. \ref{eq:SIRD_D}, and its associated variance is obtained through a parametric bootstrap assuming a multivariate normal distribution with mean and covariance equal to the estimate from the non-linear least squares technique. The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \citep{efron2012bayesian}.
